---
title: "Lab5"
author: "Felix Baez-Santiago Ryan Herren & Carter Meseck"
date: "4/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

- 0. Imports and Dataframe
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(classdata)
library(lubridate)
library(tidyverse)
library(maps)
library(ggrepel)
acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
per <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
acc2 <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
per2 <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)

makeCode <- read.csv("makeCodes.csv")
stateCode <- read.csv("stateCodes.csv")
```
- 1. Are there some days of the week where more accidents happen than the others (see FARS manual, use variable DAY_WEEK)?
```{r}
acc %>%
  mutate(recode(DAY_WEEK, "1" = "Sunday", "2" = "Monday", "3" = "Tuesday", "4" = "Wednesday", "5" = "Thursday", "6" = "Friday", "7"  = "Saturday")) -> mutAcc
head(acc)
accDays <- ggplot(acc, aes(DAY_WEEK)) + geom_bar(width=0.7) + theme_minimal()
print(accDays + labs(y="Count", x="Days of The Week", title = "Total Number of Accidents on each day of the Week"))
```

- 2. Create a data frame containing the persons who are fatally hurt in the accidents (look up variable INJ_SEV)
```{r}
fatalAccidents <- per %>%
      filter(INJ_SEV == 4)
head(fatalAccidents)
```

```

- 3. Create a data frame containing the most dangerous vehicle make in each state. The number of persons fatally hit in the vehicle make is used to assess the (non-)safety of a make. Make sure to handle the missing values appropriately. (look up variable MAKE)
```{r}
makeCode %>%
  mutate(MAKE = Codes) -> makeCode
stateCode %>%
  mutate(STATE = Code) -> stateCode
per %>%
  full_join(makeCode) -> per
per %>%
  full_join(stateCode) -> per
fatal <- per %>%
  filter(INJ_SEV == 2)
danger <- fatal %>%
  group_by(State, Make) %>%
  summarize(n = n()) %>%
  group_by(State) %>%
  filter(n == max(n))
danger
```

- 4. Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map. (Hint: Read the description for the STATE and COUNTY columns in the FARS manual. The state & county codes are Geographic Locator Codes (GLCs) from the General Services Administration’s (GSA) publication. Use readxl::read_xlsx to read in the GLCs.)
```{r}
MakeCode <- read.csv('makeCodes.csv')
StateCode <- read.csv('stateCodes.csv')
StateCode$State<- tolower(StateCode$State)
states <- map_data("state")
labelPosition <- states %>%
   group_by(region) %>%
   summarise(long= mean(long, na.rm = TRUE), lat= mean(lat, na.rm = TRUE))
labelPosition <- left_join(labelPosition, StateCode, by = c("region" = "State"))
 head(labelPosition, 10)
```
- 5. Join the accident and person table (work out which variable(s) to use)
```{r}
acc_per <- full_join(acc, per)
head(acc_per)
```

- 6. Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results and explain what you find.
```{r}
acc_per_6 <- acc_per %>%
  filter(SEX == 1 || SEX == 2) %>%
  group_by(ST_CASE) %>%
  summarise(avgSex = mean(SEX))
acc_per_6$sexStr <- ifelse(acc_per_6$avgSex == 1, 'Men', ifelse(acc_per_6$avgSex == 2, 'Women', 'Both'))
accident <- acc_per_6 %>%
  select(ST_CASE, sexStr) %>%
  inner_join(acc, by='ST_CASE')
grouped <- accident %>%
  filter(HOUR <= 24) %>%
  filter(DAY_WEEK != 9) %>%
  group_by(sexStr, HOUR, DAY_WEEK) %>%
  summarise(numAccidents = n()) %>%
  arrange(desc(numAccidents))
ggplot(grouped,aes(x=HOUR, y=numAccidents)) + 
  geom_bar(stat='identity') + 
  facet_grid(sexStr~DAY_WEEK) + 
  xlab("Hours per day") + 
  ylab("Accidents in that day")


```

- 7. Now plot a choropleth map of the number of deaths on a county level. Also explain what you find.
```{r}
county = map_data('county')
glc = readxl::read_excel('/Users/ryanherren/Desktop/DS202/ds202_lab5/FRPP_GLC_-_United_StATESaPRIL122021.xlsx')
glc$'State Code' = as.numeric(glc$'State Code')
glc$'County Code' = as.numeric(glc$'County Code')
glc$'County Name' = tolower(glc$'County Name') 
glc$'State Name' = tolower(glc$'State Name') 
q7 = acc2 %>% group_by(STATE, COUNTY) %>% summarize(num=n()) %>% select(STATE, COUNTY, num)
q7p = glc %>% left_join(q7, by=c('State Code' = 'STATE', 'County Code' = 'COUNTY')) %>% right_join(county, by=c('State Name' = 'region', 'County Name' = 'subregion'))
ggplot(q7p, aes(x=long, y=lat, fill=num)) + geom_polygon(aes(group=group)) + coord_map() + ggtitle('Fatal Accidents by County')
```
  
  I found that there might be a little correlation between being on the coast and higher fatality rates. The southern part of California has high rates, as does a coastal county of southeastern Texas. The places with the least fatalities are the ones with low populations, like Montana and the Dakotas.
  
- 8. Is summer or winter more dangerous? Does this depend on states? Explore and explain.
```{r}
seasonal_accidents = acc2 %>% 
  select(STATE, MONTH) %>% 
  left_join(glc, by=c('STATE' = 'State Code')) %>% 
  rename('state_name' = 'State Name') %>% 
  filter(state_name %in% c('montana', 'iowa', 'california', 'new york', 'florida', 'texas')) %>% 
  group_by(state_name, MONTH) %>% 
  mutate(num = n()) %>% ungroup()
seasonal_accidents %>% 
  group_by(state_name, MONTH) %>% 
  ggplot(aes(x=MONTH, fill=num)) + geom_bar() + facet_wrap(~state_name) + ylab('Number of fatal accidents') + xlab('Month') + ggtitle('Fatal Accidents')
```
  
  It doesn't really seem like much is any correlation between season and the rate of fatal accidents. In some states, the numbers go down in the summer(New York) and in some, they go down in the winters(California). It really just depends on the state for how their trends go. Iowa has the most fatal accidents in September and October, while New York has the most in June and July. I think the summers might slightly have an overall increase in deaths, and that could be attributed to people going out more in the summer because of the nice weather, as opposed to staying in their house for the winters.